# Random variables and their probability distributions {#distributions}

## Introduction

Last chapter's combinatorial probabilities are difficult to find and very problem-specific. Instead,
in this chapter we shall find easier ways to calculate probability in structured cases. The outcomes
of random experiments will be represented as values of a variable which will be random since the
outcomes are random. In so doing, we will make our life a lot
easier in calculating probabilities in many stylised situations which represent reality. 

## Random variables

### Introduction

In this section we will learn about the probability distribution of a random variable defined by
its probability function. The probability function will be called the probability mass function for
discrete random variables and the probability density function for continuous random variables.

A random variable defines a one-to-one mapping of the sample space consisting of all possible
outcomes of a random experiment to the set of real numbers. For example, I toss a coin. Assuming
the coin is fair, there are two possible equally likely outcomes: head or tail. These two outcomes
must be mapped to real numbers. For convenience, I may define the mapping which assigns the
value 1 if head turns up and 0 otherwise. Hence, we have the mapping
\[\text{Head} \rightarrow 1, \text{Tail} \rightarrow 0.\]

We can conveniently denote the random variable by $X$ which is the number of heads obtained by
tossing a single coin. The possible values of $X$ are $0$ and $1$.

You will say that this is a trivial example. Indeed it is. But it is very easy to generalise the
concept of random variables. Simply define a mapping of the outcomes of a random experiment
to the real number space. For example, I toss the coin $n$ times and count the number of heads
and denote that to be $X$. $X$ can take any real positive integer value between $0$ and
$n$. Among other examples, suppose I select a University of Southampton student at random and
measure their height. The outcome in metres will probably be a number between one metre 
and two metres. 
But I can't exactly tell which value it will be since I do not know which student will be
selected in the first place. However, when a student has been selected I can measure their height
and get a value such as $1.432$ metres.

We now introduce two notations: $X$ (or in general the capital letters $Y, Z$ etc.) to denote the
random variable, e.g. height of a randomly selected student, and the corresponding lower case letter
$x$ (or $y$, $z$) to denote a particular value, e.g. 1.432 metres. We will follow this convention throughout.
For a random variable, say $X$, we will also adopt the notation $P (X \in A)$, read
probability that $X$
belongs to $A$, instead of the previous $P \{A\}$ for any event $A$.


### Discrete and continuous random variables

If a random variable has a finite or countably infinite set of values it is called discrete. For example,
the number of Apple computer users among 20 randomly selected students, or the number of credit
cards a randomly selected person has in their wallet.
When the random variable can take any value on the real line it is called a continuous random
variable. For example, the height of a randomly selected student. A random variable can also take
a mixture of discrete and continuous values, e.g. volume of precipitation collected in a day; some
days it could be zero, on other days it could be a continuous measurement, e.g. 1.234 mm.

### Probability distribution of a random variable


Recall the first axiom of probability ($P \{S\} = 1$), which means total probability equals 1. 
Since a random variable is merely a mapping from the outcome space to the real line, the combined
probability of all possible values of the random variable must be equal to 1.
A probability distribution distributes the total probability 1 among the
possible values of the random variable.

:::{.example}
Returning to the coin-tossing experiment, if the probability of getting a head with
a coin is $p$ (and therefore the probability of getting a tail is $1 - p$), 
then the probability that $Y = 0$
is $1 - p$ and the probability that $Y = 1$ is $p$. 
This gives us the probability distribution of $Y$, and we
say that $Y$ has the probability function
\[P(Y = y) = \begin{cases} 1-p & \text{for $y = 0$} \\
p & \text{for $y = 1$.} \end{cases}\]
This is an example of the **Bernoulli distribution** with parameter $p$, the simplest discrete
distribution.
:::

:::{.example #two-coins}
Suppose we consider tossing the coin twice and again defining the random variable
$X$ to be the number of heads obtained. The values that $X$ can take are $0$, $1$ and $2$ 
with probabilities
$(1 - p)^2$ , $2p(1 - p)$ and $p^2$, respectively. Here the probability function is
\[P(X = x) = \begin{cases}
  (1 - p)^2 & \text{for $x = 0$} \\
  2p(1 - p) & \text{for $x = 1$} \\
  p^2 & \text{for $x = 2$.}
  \end{cases}\]
This is a particular case of the Binomial distribution. We will learn about it soon.
:::

In general, for a discrete random variable we define a function $f(x)$ to denote $P (X = x)$
(or $f (y)$ to denote $P (Y = y)$) and call the function $f (x)$
the probability function (pf) or probability
mass function (pmf) of the random variable $X$. Arbitrary functions cannot be a pmf since the
total probability must be 1 and all probabilities are non-negative. Hence, for $f (x)$ to be the pmf
of a random variable $X$, we require:

1. $f (x) \geq 0$ for all possible values of $x$.
2. $\sum_{\text{all $x$}} f (x) = 1$

In Example \@ref(exm:two-coins), we may rewrite the probability function in the general form
\[f(x) = \binom{2}{x} p^x (1 - p)^{2-x}, \text{for $x = 0, 1, 2$},\]
where $f (x) = 0$ for any other value of $x$.

### Continuous random variables

In many situations (both theoretical and practical) we often encounter random variables that are
inherently continuous because they are measured on a continuum (such as time, length, weight)
or can be conveniently well-approximated by considering them as continuous (such as the annual
income of adults in a population, closing share prices).

For a continuous random variable, $P (X = x)$ is defined to be zero since we assume that the
measurements are continuous and there is zero probability of observing a particular value, e.g. $1.2$.
The argument goes that a finer measuring instrument will give us an even more precise 
measurement than $1.2$ and so on. Thus for a continuous random variable we adopt the
convention that
$P (X = x) = 0$ for any particular value $x$ on the real line. 
But we define probabilities for positive
length intervals, e.g. $P (1.2 < X < 1.9)$.

For a continuous random variable $X$ we define its probability by using a continuous function
$f(x)$ which we call its probability density function, abbreviated as its pdf. With the pdf we define
probabilities as integrals, e.g.
\[P (a < X < b) = \int_a^b f (u) du,\]
which is naturally interpreted as the area under the curve $f (x)$ inside the interval $(a, b)$. 
This is demonstrated in Figure \@ref(fig:interval-probs-from-pdf).
Recall
that we do not use $f (x) = P (X = x)$ for any $x$ as by convention we set $P(X = x) = 0$.

```{r interval-probs-from-pdf, echo = FALSE, fig.cap = "The shaded area is $P (a < X < b)$ if the pdf of $X$ is the drawn curve.", fig.width = 5, fig.height = 3, out.width = "60%", fig.align = "center"}
library(ggplot2)
f <- function(x) {
    dgamma(x, shape = 3, rate = 0.5)
}

a <- 6
b <- 8

plot_data <- tibble(x = seq(a, b, length.out = 100),
                    ymax = f(x),
                    ymin = 0)

plot_data %>%
    ggplot(mapping = aes(x = x)) + 
    geom_function(fun = f) +
    geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.3) +
    xlab("x") +
    ylab("f(x)") +
    scale_x_continuous(breaks = c(0, 5, a, b, 10, 15, 20),
                       labels = c(0, 5, "a", "b", 10, 15, 20),
                       limits = c(0, 20)) +
    geom_hline(aes(yintercept = 0)) +
    theme(panel.grid.minor = element_blank(),
          panel.grid = element_blank())
```

Since we are dealing with probabilities which are always between $0$ and $1$, just any arbitrary
function f (x) cannot be a pdf of some random variable. For $f (x)$ to be a pdf, as in the discrete
case, we must have

1. $f (x) \geq 0$ for all possible values of $x$, i.e. $-\infty < x < \infty$,
2. $\int_{-\infty}^{\infty} f (u)du = 1$.

### Cumulative distribution function (cdf)

Along with the pdf we also frequently make use of another function which is called the cumulative
distribution function, abbreviated as the cdf. The cdf simply calculates the probability of the
random variable up to its argument.

For a discrete random variable $X$, the cdf is the cumulative sum
of the pmf $f(u)$ up to (and including) $u = x$. That is,
\[P (X \leq x) \equiv F (x) = \sum_{u \leq x} f (u).\]

:::{.example}
Let $X$ be the number of heads in the experiment of tossing two fair coins. Then
the probability function is
\[P (X = 0) = 1/4, \; P (X = 1) = 1/2, \; P (X = 2) = 1/4.\]

From the definition, the CDF is given by
\[F(x)= \begin{cases}0 & \text { if } x<0 \\ 1 / 4 & \text { if } 0 \leq x<1 \\ 3 / 4 & \text { if } 1 \leq x<2 \\ 1 & \text { if } x \geq 2\end{cases}\]
:::

The cdf for a discrete random variable is a step function. The jump-points are the
possible values of the random variable, and the height of a jump gives the probability of
the random variable taking that value. It is clear that the probability mass function is uniquely
determined by the cdf.

For a continuous random variable $X$, the cdf is defined as
\[P (X \leq x) \equiv F (x) = \int_{-\infty}^x f (u)du.\]

The fundamental theorem of calculus then tells us that
\[f (x) = \frac{dF (x)}{dx},\]
so for a continuous random variable the pdf is the derivative of the cdf. Also for any random
variable $X$, $P (c < X < d) = F (d) - F (c)$. Let us consider an example.

:::{.example name="Uniform distribution" #uniform-cdf}
Suppose
\[f(x)= \begin{cases}\frac{1}{b-a} & \text { if } a<x<b \\ 0 & \text { otherwise.}\end{cases}\]
In this case we say X has *uniform distribution* on the interval $(a, b)$,
which we will write as $X \sim U(a, b)$.
We now have the cdf 
\[F(x)=\int_{a}^{x} \frac{1}{b-a} du =\frac{x-a}{b-a}, \; \; a<x<b.\]
A quick check confirms that $F^{\prime}(x)=f(x)$. If $a=0, b=1$ 
then 
\[P(0.5<X<0.75)=F(0.75)-F(0.5)=0.25.\] We shall see many more examples later.
:::

## Summaries of a random variable

### Introduction

In Section \@ref(summaries), we defined various summaries of sample data $x_1 , \ldots , x_n$,
such as the mean and variance.
A random variable $X$
with either a pmf $f (x)$ or a pdf $f (x)$ may be summarised using similar measures.

### Expectation

 The mean of $X$ is called an expectation since it is a value we
can 'expect'! The expectation is defined as
\[E(X)= \begin{cases}\sum_{\text{all $x$}} x f(x) & \text { if } X \text { is discrete } \\ \int_{-\infty}^{\infty} x f(x) d x & \text { if } X \text { is continuous }\end{cases}\]
 when the sum or integral exists. 
They can't always be assumed to exist! 

Thus, roughly speaking:
the expected value is either sum or integral of value times probability.
We use the $E(\cdot)$ notation to denote expectation. The argument is in upper case since it is the expected value of the random variable which is denoted by an upper case letter. We often use the Greek letter $\mu$ to denote $E(X)$.

:::{.example text="Die throwing"}
Consider the fair-die tossing experiment, with each of the six sides having a probability of $1 / 6$ of landing face up. Let $X$ be the number on the up-face of the die. Then
\[E(X)=\sum_{x=1}^{6} x P(X=x)=\sum_{x=1}^{6} x / 6=3.5.\]
:::

:::{.example text="Uniform distribution" #uniform-mean}
Suppose $X \sim U(a, b)$, with pdf $f(x)=\frac{1}{b-a}, \, a<x<b.$ Then
\begin{align*}
E(X) &=\int_{-\infty}^{\infty} x f(x) d x \\
&=\int_{a}^{b} \frac{x}{b-a} d x \\
&=\frac{b^{2}-a^{2}}{2(b-a)}=\frac{b+a}{2},
\end{align*}
the mid-point of the interval $(a, b)$.
:::

If $Y=g(X)$ for any function $g(\cdot)$, then $Y$ is a random variable as well. To find $E(Y)$ we simply use the value times probability rule, i.e. the expected value of $Y$ is either sum or integral of its value, $g(x)$, times probability $f(x)$:
\[E(Y)=E(g(X))=\begin{cases}
\sum_{\text{all $x$}} g(x) f(x) & \text { if $X$  is discrete, } \\
\int_{-\infty}^{\infty} g(x) f(x) d x & \text { if $X$ is continuous.}
 \end{cases}\]
 
For example, if $X$ is continuous, then $E(X^2) = \int_{-\infty}^{\infty} x^2 f(x) dx$.
We prove one important property
of expectation, namely expectation is a linear operator.

::: {.theorem}
Suppose $Y = g(X) = aX +b$; then $E(Y ) = aE(X)+b$.
:::

::: {.proof}
The proof is given for the continuous case. In the discrete case replace
integral ($\int$) by summation ($\sum$).
\begin{align*}
E(Y ) &= \int_{-\infty}^\infty (ax + b) f(x)dx \\
&= a \int_{-\infty}^\infty x f (x)dx + b \int_{-\infty}^\infty f (x)dx \\
&= aE(X) + b,
\end{align*}
using the total probability is 1 property
($\int_{-\infty}^\infty f (x)dx = 1$) in the last integral.
:::

 This is very convenient, e.g. suppose $E(X) = 5$ and
$Y = -2X + 549$ then $E(Y) = 539$.

### Variance

The variance measures the variability of a random variable and is defined by
\[\operatorname{Var}(X)=E(X-\mu)^{2}= \begin{cases}
\sum_{\text {all } x}(x-\mu)^{2} f(x) & \text { if $X$ is discrete} \\
\int_{-\infty}^{\infty}(x-\mu)^{2} f(x) dx & \text { if $X$ is continuous,}
\end{cases}\]
where $\mu=E(X)$, and when the sum or integral exists. 
When the variance exists, 
it is the expectation of $(X-\mu)^{2}$ where $\mu$ is the mean of $X$.
We now derive an easy formula to calculate the variance:

::: {.theorem}
\[\operatorname{Var}(X)=E(X-\mu)^{2}=E\left(X^{2}\right)-\mu^{2}.\]
:::

::: {.proof}
We have
\begin{align*}
\operatorname{Var}(X) &=E(X-\mu)^{2} \\
&=E\left(X^{2}-2 X \mu+\mu^{2}\right) \\
&=E\left(X^{2}\right)-2 \mu E(X)+\mu^{2} \\
&=E\left(X^{2}\right)-2 \mu \mu+\mu^{2} \\
&=E\left(X^{2}\right)-\mu^{2}.
\end{align*}
:::

Thus the variance of a random variable is the expected value of its square
minus the square of its expected value.

We usually denote the variance by $\sigma^2$. 
The square is there to emphasise that the variance of any
random variable is always non-negative. When can the variance be zero? 
When there is no variation
at all in the random variable, i.e. it takes only a single value µ with probability 1.
Hence, there is
nothing random about the random variable --- we can predict its outcome with certainty.

The square root of the variance is called the *standard deviation* of the
random variable.

:::{.example name="Uniform distribution" #uniform-variance}
Suppose $X \sim U(a, b)$, with pdf $f(x)=\frac{1}{b-a}, \, a<x<b.$  Then
\begin{align*}
E\left(X^{2}\right) &=\int_{a}^{b} \frac{x^{2}}{b-a} d x \\
&=\frac{b^{3}-a^{3}}{3(b-a)} \\
&=\frac{b^{2}+a b+a^{2}}{3}.
\end{align*}
Hence
\[\operatorname{Var}(X)=\frac{b^{2}+a b+a^{2}}{3}-\left(\frac{b+a}{2}\right)^{2}=\frac{(b-a)^{2}}{12},\]
after simplification.
:::

We prove one important property of the variance.

::: {.theorem}
Suppose $Y=a X+b$ then $\operatorname{Var}(Y)=a^{2} \operatorname{Var}(X)$
:::

::: {.proof}
Write $\mu = E(X)$. Then $E(Y) = a\mu + b$ and
\begin{align*}
\operatorname{Var}(Y) &=E\left[(Y-E(Y))^{2}\right] \\
&=E\left[(aX + b - a\mu - b)^2\right] \\
&=E\left[a^2(X -\mu)^2\right] \\
&=a^2 E\left[(X -\mu)^2\right] \\
&=a^{2} \operatorname{Var}(X)
\end{align*}
:::

This is a very useful result, e.g. suppose $\operatorname{Var}(X)=25$ and $Y=-X+5,000,000$; then $\operatorname{Var}(Y)=$ $\operatorname{Var}(X)=25$ and the standard deviation, $\sigma=5$. In words a location shift, $b$, does not change variance but a multiplicative constant, $a$ say, gets squared in variance, $a^{2}$.

### Quantiles

For a given $0<p<1$, a $p$th *quantile* (or $100 p$ *percentile*) of the random variable $X$ with 
cdf $F(x)$
is defined to be a value $q$ for which $F(q)=p$. If the cdf is invertible, we have
$q = F^{-1}(p)$.

The 50th percentile is called the *median*. The 25th and 75th percentiles are called the *quartiles*.

:::{.example name="Uniform distribution" #uniform-quantiles}
Suppose $X \sim U(a, b)$, with pdf $f(x)=\frac{1}{b-a}, \, a<x<b.$
We have shown in Example \@ref(exm:uniform-cdf) that the cdf is \[F(x)=\frac{x-a}{b-a}, \; a<x<b.\]
So for a given $p$, $F(q)=p$ implies \[q=a+p(b-a).\]
The median of $X$ is $\frac{b+a}{2}$
and the quartiles are $\frac{b + 3a}{4}$ and $\frac{3b + a}{4}$.
:::

## Standard discrete distributions

### Bernoulli distribution

A set of independent trials, where each trial
has only two possible outcomes, conveniently called success (S) and failure (F), and the probability
of success is the same in each trial are called a set of *Bernoulli trials*. 

Suppose that we conduct one Bernoulli trial, where we get a success $(S)$ or failure $(F)$ with probabilities $P\{S\}=p$ and $P\{F\}=1-p$ respectively. 
Let $X$ be an indicator of success:
\[X = \begin{cases} 1 & \text{if $S$} \\
0 & \text{if $F$.}
\end{cases}\]
Then $X$ has Bernoulli distribution with parameter $p$, written as
$X \sim \operatorname{Bernoulli}(p)$.

The Bernoulli distribution has pmf \[f(x)=p^{x}(1-p)^{1-x}, x=0,1.\]
Hence \[E(X)=0 \cdot(1-p)+1 \cdot p=p,\]
\[E\left(X^{2}\right)=0^{2} \cdot(1-p)+1^{2} \cdot p=p\] and
\[\operatorname{Var}(X)=E\left(X^{2}\right)-(E(X))^{2}=p-p^{2}=p(1-p).\]

### Binomial distribution

#### Introduction and definition
Suppose that we have a sequence of $n$ Bernoulli trials such that we get a success with probability $p$. Let $X$ be the number of successes in the $n$ trials. Then $X$ has binomial distribution with parameters $n$ and $p$, written as
$X \sim \operatorname{Bin}(n, p)$.

An outcome of the experiment (of carrying out $n$ such independent trials) is represented by a sequence of $S$'s and $F$'s (such as $S S \ldots F S \ldots S F)$ 
that comprises $x$ $S$ 's, and $(n-x)$ $F$ 's.
The probability associated with this outcome is
\[P\{S S \ldots F S \ldots S F\}=p p \cdots(1-p) p \cdots p(1-p)=p^{x}(1-p)^{n-x}.\]
For this sequence, $X=x$, but there are many other sequences which will also give $X=x$. 
In fact there are $\binom{n}{x}$ such sequences. Hence
\[P(X=x)= \binom{n}{x} p^{x}(1-p)^{n-x}, x=0,1, \ldots, n .\]
This is the pmf of the binomial distribution with parameters $n$ and $p$

How can we guarantee that $\sum_{x=0}^{n} P(X=x)=1$? This guarantee is provided by the binomial theorem:

::: {.theorem name="Binomial theorem" #bin-thm}
For any positive integer $n$ and
 real numbers $a$ and $b$,
\[(a+b)^{n}=b^{n}+ \binom{n}{1} a b^{n-1}+\cdots+\binom{n}{x} a^x b^{n-x}+\cdots +a^{n}.\]

:::

To prove $\sum_{x=0}^{n} P(X=x)=1$, i.e. $\sum_{x=0}^{n}\binom{n}{x} p^{x}(1-p)^{n-x}=1$, 
choose $a=p$ and $b=1-p$ in the binomial theorem.


:::{.example #widgets}
Suppose that widgets are manufactured in a mass production process with $1\%$
defective. The widgets are packaged in bags of 10 with a money-back guarantee if more than 1
widget per bag is defective. For what proportion of bags would the company have to provide a
refund?

First, we find the probability that a randomly selected bag has at most 1 defective
widget. Let $X$ be the number of defective widgets in a bag, then 
$X \sim \operatorname{Bin}(n = 10, p = 0.01).$
So this
probability is equal to
\[P (X = 0) + P (X = 1) = (0.99)^{10} + 10(0.01)^1 (0.99)^9 = 0.9957.\]
Hence the probability that a refund is required is $1 - 0.9957 = 0.0043$, 
i.e. only just over 4 in 1000
bags will incur the refund on average.
:::

:::{.example}
A binomial random variable can also be described using the urn model. Suppose
we have an urn (population) containing $N$ individuals, a proportion $p$ of which are of type $S$ and a
proportion $1 - p$ of type $F$. If we select a sample of $n$ individuals at random with replacement,
then the number, $X$, of type $S$ individuals in the sample follows the binomial distribution with
parameters $n$ and $p$.
:::

#### Using R to calculate probabilities

Probabilities under all the standard distributions have been calculated in R and will be used
throughout MATH1024. You will not be required to use any tables. 
For the binomial distribution
the command 
```{r, eval = FALSE}
dbinom(x=3, size=5, prob=0.34)
```
calculates the pmf of $\operatorname{Bin}(n = 5, p = 0.34)$ at $x=3$,
with value $P (X = 3) = \binom{5}{3} (0.34)^3 (1 - 0.34)^{5-3}$. 
The command `pbinom` returns the cdf or the probability up to
and including the argument. Thus 
```{r, eval = FALSE}
pbinom(q=3, size=5, prob=0.34)
```
will return the value of
$P (X \leq 3)$ when $X \sim \operatorname{Bin}(n = 5, p = 0.34)$. 
As a check, in Example \@ref(exm:widgets), we may compute the probability
that a randomly selected bag has at most 1 defective
widget with the command
```{r}
pbinom(q=1, size=10, prob=0.01)
```
which matches our earlier calculations.

#### Expectation {#sec:bin-mean}

Let $X \sim \operatorname{Bin}(n, p)$. We have
\[E(X)=\sum_{x=0}^{n} x P(X=x)=\sum_{x=0}^{n} x \binom{n}{x} p^{x}(1-p)^{n-x}.\]
Below we prove that $E(X)=n p$. Recall that $k !=k(k-1) !$ for any $k>0$.
\begin{align*}
E(X) &=\sum_{x=0}^{n} x \binom{n}{x} p^{x}(1-p)^{n-x} \\
&=\sum_{x=1}^{n} x \frac{n!}{x !(n-x)!} p^{x}(1-p)^{n-x} \\
&=\sum_{x=1}^{n} \frac{n !}{(x-1) !(n-x) !} p^{x}(1-p)^{n-x} \\
&=n p \sum_{x=1}^{n} \frac{(n-1) !}{(x-1) !(n-1-x+1) !} p^{x-1}(1-p)^{n-1-x+1} \\
&=n p \sum_{y=0}^{n-1} \frac{(n-1) !}{y!(n-1-y)!} p^{y}(1-p)^{n-1-y} \\
&=n p \sum_{y=0}^{n-1} \binom{n-1}{y} p^{y}(1-p)^{n-1-y} \\
&=n p
\end{align*}
where we used the substitution $y=x-1$ and then
conclude the last sum equals one as it is the sum of all
probabilities in the $\operatorname{Bin}(n-1, p)$ distribution.


#### Variance {#sec:bin-var}

Let $X \sim \operatorname{Bin}(n, p)$. Then $\operatorname{Var}(X)=n p(1-p)$. 
It is difficult to find $E\left(X^{2}\right)$ directly, but the factorial structure
allows us to find $E[X(X-1)]$. Recall that $k !=k(k-1)(k-2) !$ for any $k>1$.
\begin{align*}
E[(X(X-1)]&=\sum_{x=0}^{n} x(x-1) \binom{n}{x} p^{x}(1-p)^{n-x} \\
&=\sum_{x=2}^{n} x(x-1) \frac{n !}{x !(n-x) !} p^{x}(1-p)^{n-x} \\
&=\sum_{x=2}^{n} \frac{n !}{(x-2) !(n-x) !} p^{x}(1-p)^{n-x} \\
&=n(n-1) p^{2} \sum_{x=2}^{n} \frac{(n-2) !}{(x-2) !(n-2-x+2) !} p^{x-2}(1-p)^{n-2-x+2} \\
&=n(n-1) p^{2} \sum_{y=0}^{n-2} \frac{(n-2) !}{(y) !(n-2-y) !} p^{y}(1-p)^{n-2-y} \\
&=n(n-1) p^2
\end{align*}
where we used the substitution $y=x-2$ and then
conclude the last sum equals one as it is the sum of all
probabilities in the $\operatorname{Bin}(n-2, p)$ distribution.
Now, $E\left(X^{2}\right)=E[X(X-1)]+E(X)=n(n-1) p^{2}+n p$. Hence,
\[\operatorname{Var}(X)=E\left(X^{2}\right)-(E(X))^{2}=n(n-1) p^{2}+n p-(n p)^{2}=n p(1-p).\]
It is illuminating to see these direct proofs. Later on we shall apply statistical theory to directly prove these.

### Geometric distribution

#### Introduction and definition
Suppose that we have the same situation as for the binomial distribution but we consider a different random variable $X$, which is defined as the number of trials that lead to the first success. The outcomes for this experiment are:
\[\begin{array}{rll}
S & X=1, & P(X=1)=p \\
F S & X=2, & P(X=2)=(1-p) p \\
F F S & X=3, & P(X=3)=(1-p)^{2} p \\
F F F S & X=4, & P(X=4)=(1-p)^{3} p \\
\vdots & \vdots &
\end{array}\]
In general we have
\[P(X=x)=(1-p)^{x-1} p, \; x=1,2, \ldots\]
This is called the **geometric** distribution, and it has a (countably) infinite domain starting at $1$ rather than $0$. We write $X \sim \operatorname{Geo}(p)$.

:::{.example}
In a board game that uses a single fair die, a player cannot start until they have
rolled a six. Let $X$ be the number of rolls needed until they get a six. Then $X$ is a Geometric
random variable with success probability $p = 1/6$.
:::


In order to check the probability function sums to one, we will need
to use the general result on the geometric series:

::: {.theorem name="Geometric series"}
For any real numbers $a$ and $r$ such that $|r| < 1$,
\[\sum_{k=0}^\infty a r^k = \frac{a}{1-r}.\]
:::

We now check the probability function sums to one:
\begin{align*}
\sum_{x=1}^{\infty} P(X=x) &=\sum_{x=1}^{\infty}(1-p)^{x-1} p \\
&=\sum_{y=0}^{\infty}(1-p)^{y} p \quad \text{(substitute  $y=x-1$)} \\
&= \frac{p}{1-(1-p)} \quad \text{(geometric series, $a=p$, $r=1-p$)} \\
&=1
\end{align*}
We can also find the probability that $X>k$ for some given positive integer $k$ :
\begin{align*}
\sum_{x=k+1}^{\infty} P(X=x) &=\sum_{x=k+1}^{\infty}(1-p)^{x-1} p \\
&=p\left[(1-p)^{k+1-1}+(1-p)^{k+2-1}+(1-p)^{k+3-1}+\ldots\right]\\
&=p(1-p)^{k} \sum_{y=0}^{\infty}(1-p)^{y} \\
&=(1-p)^{k}
\end{align*}



#### Memoryless property 

Let $X$ follow the geometric distribution and suppose that $s$ and $k$ are positive integers. We then have
\[P(X>s+k \mid X>k)=P(X>s).\]

The proof is given below. In practice this means that the random variable does not remember its age (denoted by $k$) to determine how long more (denoted by $s$) it will survive! The proof below uses the definition of conditional probability
\[P\{A \mid B\}=\frac{P\{A \cap B\}}{P\{B\}}.\]
Now the proof,
\begin{align*}
P(X>s+k \mid X>k) &=\frac{P(X>s+k, X>k)}{P(X>k)} \\
&=\frac{P(X>s+k)}{P(X>k)} \\
&=\frac{(1-p)^{s+k}}{(1-p)^{k}} \\
&=(1-p)^{s},
\end{align*}
which does not depend on $k$. Note that the event $X>s+k$ and $X>k$ implies and is implied by $X>s+k$ since $s>0$.


#### Expectation and variance
Let $X \sim \operatorname{Geo}(p)$. We can show that $E(X)=1/p$ using the negative binomial series:

::: {.theorem name="Negative binomial series"}
For any positive integer $n$ and real number $x$ such that $|x|<1$
\[(1-x)^{-n}=1+n x+\frac{1}{2} n(n+1) x^{2}+\frac{1}{6} n(n+1)(n+2) x^{3}+\cdots+\frac{n(n+1)(n+2) \cdots(n+k-1)}{k !} x^{k}+\cdots\]
:::

We have
\begin{align*}
E(X) &=\sum_{x=1}^{\infty} x P(X=x) \\
&=\sum_{x=1}^{\infty} x p(1-p)^{x-1} \\
&=p\left[1+2(1-p)+3(1-p)^{2}+4(1-p)^{3}+\ldots\right]
\end{align*}
The series in the square brackets is the negative binomial series with $n=2$ and $x=1-p$.
Thus $E(X)=p(1-1+p)^{-2}=1 / p$. It can be shown that $\operatorname{Var}(X)=(1-p) / p^{2}$ using negative binomial series. But this is more complicated and is not required here. 
The second-year module MATH2011 will provide an alternative proof.

### Hypergeometric distribution

Suppose we have an urn (population) containing $N$ individuals, a proportion $p$ of which are of type $S$ and a proportion $1-p$ of type $F$. If we select a sample of $n$ individuals at random without replacement, then the number, $X$, of type $S$ individuals in the sample has the hypergeometric distribution,
with pmf
\[P(X=x)=\frac{\binom{Np}{x} \binom{N(1-p)}{n-x}}{\binom{N}{n}}, \quad x=0,1, \ldots, n,\]
assuming that $x \leq N p$ and $n-x \leq N(1-p)$ so that the above combinations are well defined. 
The mean and variance of the hypergeometric distribution are given by
\[E(X)=n p, \quad \operatorname{Var}(X)=n p q \frac{N-n}{N-1}.\]

### Negative binomial distribution

Still in the Bernoulli trials set-up, we define the random variable $X$ to be the total number
of trials
until the $r$th success occurs, where $r$ is a given postive integer. This is known as the negative
binomial distribution with parameters $p$ and $r$.
[Note: if $r = 1$, the negative binomial distribution is just the geometric distribution.]
Firstly we need to identify the possible values of $X$. Possible values for $X$ are $x = r, r + 1, r +
2, \ldots$. The probability mass function is
\begin{align*}
P (X = x) &= \binom{x-1}{r-1}
p^{r-1} (1 - p)^{(x-1)-(r-1)} \times p \\
&=\binom{x-1}{r-1}
p^r (1 - p)^{x-r}, \quad x = r, r + 1, \ldots
\end{align*}


:::{.example} 
A man plays roulette, betting on red each time. He decides to keep playing until
he achieves his second win. The success probability for each game is $18/37$ and the results of games
are independent. Let $X$ be the number of games played until he gets his second win. Then $X$ is
a negative binomial random variable with $r = 2$ and p = $18/37$.

What is the probability he plays
more than $3$ games? We have
\[P (X > 3) = 1 - P(X = 2) - P(X = 3)  =  1 - p^2 - 2 p^2 (1-p) = 0.520.\]
:::

Derivation of the mean and variance of the negative binomial distribution involves complicated negative binomial series and will be skipped for now, but will be proved in Section \ref{sec:sum-rvs}.
For
completeness we note down the mean and variance:
\[E(X) = \frac{r}{p}, \quad \operatorname{Var}(X) = r \, \frac{1-p}{p^2}.\]
Thus when r = 1, the mean and variance of the negative binomial distribution are equal to those
of the geometric distribution.

### Poisson distribution

#### Introduction and definition

The Poisson distribution can be obtained as the limit of the binomial distribution with parameters $n$ and $p$ when $n \rightarrow \infty$ and $p \rightarrow 0$ simultaneously, but the product $\lambda=n p$ remains finite. In practice this means that the Poisson distribution counts rare events (since $p \rightarrow 0$ ) in an infinite population (since $n \rightarrow \infty$ ). Theoretically, a random variable following the Poisson distribution can take any integer value from 0 to $\infty$. Examples of the Poisson distribution include: the number of breast cancer patients in Southampton; the number of text messages sent (or received) per day by a randomly selected first-year student; the number of credit cards a randomly selected person has in their wallet.

Let us find the pmf of the Poisson distribution as the limit of the pmf of the binomial distribution. Recall that if $X \sim \operatorname{Bin}(n, p)$ then $P(X=x)=\left(\begin{array}{l}n \\ x\end{array}\right) p^{x}(1-p)^{n-x}$. Now:
\begin{align*}
P(X=x) &=\binom{n}{x} p^{x}(1-p)^{n-x} \\
&=\binom{n}{x} \frac{n^{n}}{n^{n}} p^{x}(1-p)^{n-x} \\
&=\frac{n(n-1) \cdots(n-x+1)}{n^{x} x !}(n p)^{x}(n(1-p))^{n-x} \frac{1}{n^{n-x}} \\
&=\frac{n}{n} \frac{(n-1)}{n} \cdots \frac{(n-x+1)}{n} \frac{\lambda^{x}}{x !}\left(1-\frac{\lambda}{n}\right)^{n-x}\\
&=\frac{n}{n} \frac{(n-1)}{n} \cdots \frac{(n-x+1)}{n} \frac{\lambda^{x}}{x !}\left(1-\frac{\lambda}{n}\right)^{n}\left(1-\frac{\lambda}{n}\right)^{-x}.
\end{align*}
Now it is easy to see that the above tends to
\[e^{-\lambda} \frac{\lambda^{x}}{x!}\]
as $n \rightarrow \infty$ for any fixed value of $x$ in the range $0,1,2, \ldots .$ Note that we have used the exponential limit:
\[e^{-\lambda}=\lim _{n \rightarrow \infty}\left(1-\frac{\lambda}{n}\right)^{n},\]
and
\[ \lim _{n \rightarrow \infty}\left(1-\frac{\lambda}{n}\right)^{-x}=1\]
and
\[ \lim _{n \rightarrow \infty} \frac{n}{n} \frac{(n-1)}{n} \cdots \frac{(n-x+1)}{n}=1.\]
A random variable $X$ has the Poisson distribution with parameter $\lambda$ if it has the pmf:
\[P(X=x)=e^{-\lambda} \frac{\lambda^{x}}{x !}, \quad x=0,1,2, \ldots\]
We write $X \sim \operatorname{Poisson}(\lambda)$. It is easy to show $\sum_{x=0}^{\infty} P(X=x)=1$, i.e. $\sum_{x=0}^{\infty} e^{-\lambda} \frac{\lambda^{x}}{x !}=1$. The identity you need is simply the expansion of $e^{\lambda}$.

#### Expectation

Let $X \sim \operatorname{Poisson}(\lambda)$. Then
\begin{align*}
E(X) &=\sum_{x=0}^{\infty} x P(X=x) \\
&=\sum_{x=0}^{\infty} x e^{-\lambda} \frac{\lambda^{x}}{x!} \\
&=e^{-\lambda} \sum_{x=1}^{\infty} x \frac{\lambda^{x}}{x !} \\
&=e^{-\lambda} \sum_{x=1}^{\infty} \frac{\lambda \cdot \lambda^{(x-1)}}{(x-1) !} \\
&=\lambda e^{-\lambda} \sum_{x=1}^{\infty} \frac{\lambda^{(x-1)}}{(x-1) !} \\
&=\lambda e^{-\lambda} \sum_{y=0}^{\infty} \frac{\lambda^{y}}{y !} \quad (y=x-1) \\
&=\lambda e^{-\lambda} e^{\lambda} \quad \text {using the expansion of $e^{\lambda}$}  \\
&=\lambda .
\end{align*}

#### Variance
Let $X \sim \operatorname{Poisson}(\lambda)$. Then
\begin{align*}
E[X(X-1)] &=\sum_{x=0}^{\infty} x(x-1) P(X=x) \\
&=\sum_{x=0}^{\infty} x(x-1) e^{-\lambda} \frac{\lambda^{x}}{x!} \\
&=e^{-\lambda} \sum_{x=2}^{\infty} x(x-1) \frac{\lambda^{x}}{x !} \\
&=e^{-\lambda} \sum_{x=2}^{\infty} \lambda^{2} \frac{\lambda^{x-2}}{(x-2) !} \\
&=\lambda^{2} e^{-\lambda} \sum_{y=0}^{\infty} \frac{\lambda^{y}}{y !} \quad (y=x-2) \\
&=\lambda^{2} e^{-\lambda} e^{\lambda}=\lambda^{2} \quad \text {using the expansion of $e^{\lambda}$.} 
\end{align*}
Now, $E\left(X^{2}\right)=E[X(X-1)]+E(X)=\lambda^{2}+\lambda$. Hence,
\[\operatorname{Var}(X)=E\left(X^{2}\right)-(E(X))^{2}=\lambda^{2}+\lambda-\lambda^{2}=\lambda.\]
Hence, the mean and variance are the same for the Poisson distribution.

#### Using R to calculate probabilities

For the Poisson distribution the command 
```{r, eval = FALSE}
dpois(x=3, lambda=5)
```
calculates the pmf of $\operatorname{Poisson}(\lambda = 5)$ at $x = 3$. 
That is, the command will return the value $P (X = 3) = e^{-5} \frac{5^3}{3!}$. The command `ppois`
returns the cdf or the probability up to and including the argument. Thus 
```{r, eval = FALSE}
ppois(q=3, lambda=5)
```
will return the value of $P (X \leq 3)$ when $X \sim \operatorname{Poisson}(\lambda = 5)$.

## Standard continuous distributions

### Uniform distribution

#### Definition and properties

A continuous random variable $X$ is said to follow the uniform distribution if its pdf is of the form:
\[f(x)= \begin{cases} \frac{1}{b-a} & \text {if $a < x < b$} \\ 0 & \text {otherwise}\end{cases}\]
where $a < b$ are parameters. We write $X \sim U(a, b)$.

We have already derived various properties the uniform distribution:

- **cumulative distribution function**:
\[F(x) =\frac{x-a}{b-a}, \; a<x<b\] from Example \@ref(exm:uniform-cdf).
- **expectation**:
\[E(X) = \frac{b+a}{2} \] from Example \@ref(exm:uniform-mean).
- **variance**:
\[\operatorname{Var}(x) = \frac{(b-a)^{2}}{12}\] from Example \@ref(exm:uniform-variance).
- **quantiles**: The $p$th quantile is $a+p(b-a)$ from Example \@ref(exm:uniform-quantiles).
The median is $\frac{b+a}{2}$.

#### Using R to calculate probabilities

For $X \sim U(a = -1, b = 1)$, the command 
```{r, eval = FALSE}
dunif(x = 0.5, min = -1, max = 1)
```
calculates the pdf at $x=0.5$. We specify $a$ with the `min` argument
and $b$ with `max` argument.
The command `punif` returns the cdf or the probability up to and including the argument. 
Thus 
```{r, eval = FALSE}
punif(q = 0.5, min = -1, max = 1)
```
will return the value of $P(X \leq 0.5)$.

The command `qunif` can be used to calculate quantiles. Thus
```{r, eval = FALSE}
qunif(p = 0.5, min = -1, max = 1)
```
finds the median (the $0.5$ quantile).

### Exponential distribution

#### Introduction and definition

A continuous random variable $X$ is said to follow the exponential distribution if its pdf is of the form:
\[f(x)= \begin{cases}\theta e^{-\theta x} & \text { if $x>0$} \\ 0 & \text { if $x \leq 0$}\end{cases}\]
where $\theta>0$ is a parameter. We write $X \sim \operatorname{Exponential}(\theta)$. 
The distribution only resides in the positive half of the real line, and the tail goes 
down to zero exponentially as $x \rightarrow \infty$. The rate at which that happens
is the parameter $\theta$. Hence $\theta$ is known as the rate parameter.

It is easy to prove that $\int_{0}^{\infty} f(x) d x=1$. This is left as an exercise. To find the mean and variance of the distribution we need to introduce the *gamma function*:

The gamma function $\Gamma(.)$ is defined for any positive number $a$ as
\[\Gamma(a)=\int_{0}^{\infty} x^{a-1} e^{-x} d x\]

We have the following facts:
\[\Gamma\left(\frac{1}{2}\right)=\sqrt{\pi} ; \quad \Gamma(1)=1 ; \quad \Gamma(a)=(a-1) \Gamma(a-1) \text { if $a>1$}\]
These last two facts imply that $\Gamma(k)=(k-1)!$ when $k$ is a positive integer. 
Find $\Gamma\left(\frac{3}{2}\right)$.

#### Expectation and variance
By definition,
\begin{align*}
E(X) &=\int_{-\infty}^{\infty} x f(x) d x \\
&=\int_{0}^{\infty} x \theta e^{-\theta x} d x \\
&=\int_{0}^{\infty} y e^{-y} \frac{d y}{\theta} \quad \text {(substitute $y=\theta x$)}         \\
&=\frac{1}{\theta} \int_{0}^{\infty} y^{2-1} e^{-y} d y \\
&=\frac{1}{\theta} \Gamma(2) \\
&=\frac{1}{\theta} \quad \text { since } \Gamma(2)=1 !=1 .
\end{align*}
Now,
\begin{align*}
E\left(X^{2}\right) &=\int_{-\infty}^{\infty} x^{2} f(x) d x \\
&=\int_{0}^{\infty} x^{2} \theta e^{-\theta x} d x \\
&=\theta \int_{0}^{\infty}\left(\frac{y}{\theta}\right)^{2} e^{-y} \frac{d y}{\theta} \quad \text {(substitute $y=\theta x$)} \\
&=\frac{1}{\theta^{2}} \int_{0}^{\infty} y^{3-1} e^{-y} d y \\
&=\frac{1}{\theta^{2}} \Gamma(3) \\
&=\frac{2}{\theta^{2}} \quad \text { since } \Gamma(3)=2 !=2  ,
\end{align*}
and so 
\[\operatorname{Var}(X)=E\left(X^{2}\right)-[E(X)]^{2}=2 / \theta^{2}-1 / \theta^{2}=1 / \theta^{2}.\]
For this random variable the mean is equal to the standard deviation.

#### Using R to calculate probabilities
For $X \sim \operatorname{Exponential}(\theta=0.5)$, the command 
```{r, eval = FALSE}
dexp(x = 3, rate = 1/2)
```
calculates the pdf at $x=3$. The rate parameter to be supplied is the $\theta$ parameter here.
The command `pexp` returns the cdf or the probability up to and including the argument. 
Thus 
```{r, eval = FALSE}
pexp(q = 3, rate = 1/2)
```
will return the value of $P(X \leq 3)$.

#### Cumulative distribution function and quantiles

It is easy to find the cdf of the exponential distribution. For $x>0$,
\[F(x)=P(X \leq x)=\int_{0}^{x} \theta e^{-\theta u} d u=1-e^{-\theta x}.\]
We have $F(0)=0$ and $F(x) \rightarrow 1$ when $x \rightarrow \infty$ and $F(x)$ is non-decreasing in $x$. The cdf can be used to solve many problems. A few examples follow.

:::{.example name="Mobile phone" #mobile-phone}
Suppose that the lifetime of a phone (e.g. the time until the phone does not function even after repairs), denoted by $X$, manufactured by the company A Pale, is exponentially distributed with mean 550 days.
1. Find the probability that a randomly selected phone will still function after two years, i.e. $X>730$ ? (Assume there is no leap year in the two years.)
2. What are the times by which we expect $25 \%, 50 \%, 75 \%$ and $90 \%$ of the manufactured phones to have failed?

Here the mean $1 / \theta=550$. Hence $\theta=1 / 550$ is the rate parameter. The solution to the first problem is
\[P(X>730)=1-P(X \leq 730)=1-\left(1-e^{-730 / 550}\right)=e^{-730 / 550}=0.2652.\]
Alternatively, we can do the calculation in R:
```{r}
1 - pexp(q = 730, rate = 1 / 550)
```
:::

For the second problem we are given the probabilities of failure ($0.25$, $0.50$, etc.). We will have to invert the probabilities to find the value of the random variable. In other words, we
will have to find a $q$ such that $F(q)=p$, where $p$ is the given probability:
the $p$th quantile of $X$.

The cdf of the exponential distribution is $F(q)=1-e^{-\theta q}$, 
so to find the pth quantile we must solve $p = 1-e^{-\theta q}$ for $q$.
\begin{align*}
p &=1-e^{-\theta q} \\
\Rightarrow e^{-\theta q} &=1-p \\
\Rightarrow-\theta q &=\log (1-p) \\
\Rightarrow q &=\frac{-\log (1-p)}{\theta}.
\end{align*}
In Example \@ref(exm:mobile-phone), $\theta=1 / 550$, so we have
\[q=-550 \times \log (1-p) = \begin{cases}
158 & \text{for $p = 0.25$} \\
381 & \text{for $p = 0.50$} \\
762 & \text{for $p = 0.75$} \\
1266 & \text{for $p = 0.90$.}
  \end{cases}
  \]
which gives the time in days until we expect $25 \%, 50 \%, 75 \%$ and $90 \%$ of the manufactured phones to have failed.

In R you can find these values by
```{r, eval = FALSE}
qexp(p = 0.25, rate = 1/550)
qexp(p = 0.50, rate = 1/550)
``` 
and so on.
The function
```{r, eval = FALSE}
qexp(p, rate)
```
calculates the $p$th quantile of the exponential
distribution with parameter `rate`.

:::{.example name="Survival function"}
The exponential distribution is sometimes used to model
the survival times in different experiments. For example, an exponential random variable 
$T$ may be assumed to model the number of days a cancer patient survives after chemotherapy. 
In such a situation, the function $S(t) = 1 - F (t) = e^{-\theta t}$ is called the survival function.
See Figure \@ref(fig:survival-probs) for an example plot.

```{r, survival-probs, echo = FALSE, fig.cap = "$S(t)$ for $\\theta = 0.2, 0.5, 1$.", fig.width = 5, fig.height = 3, out.width = "60%", fig.align = "center"}
plot_data <- crossing(t = seq(0, 20, length.out = 100),
                      theta = c(1, 0.5, 0.2)) %>%
    mutate(survival_prob = pexp(t, rate = theta, lower.tail = FALSE))

plot_data %>%
    mutate(theta = as.character(theta)) %>%
    ggplot(aes(x = t, y = survival_prob, linetype = theta)) +
    geom_line() +
    ylab(label = "S(t)")
           
```
    
Assuming the mean survival time to be $100$ days for a fatal late detected cancer, we can expect
that half of the patients survive $69.3$ days after chemo since 
```{r}
qexp(0.50, rate=1/100)
```
You will learn more about this in a third-year module, MATH3085: Survival Models, 
which is important in the actuarial profession.

:::
                                   
#### Memoryless property 

Like the geometric distribution, the exponential distribution also has the memoryless property. 
In simple terms, it means that the probability that the system will survive an additional period 
$s>0$ given that it has survived up to time $t$ is the same as the probability that the system 
survives the period $s$ to begin with. That is, it forgets that it has survived up to a particular
time when it is thinking of its future remaining life time.

The proof is exactly as in the case of the geometric distribution, reproduced below. 
Recall the definition of conditional probability:
\[P\{A \mid B\}=\frac{P\{A \cap B\}}{P\{B\}}.\]
Now the proof,
\begin{align*}
P(X>s+t \mid X>t) &=\frac{P(X>s+t, X>t)}{P(X>t)} \\
&=\frac{P(X>s+t)}{P(X>t)} \\
&=\frac{e^{-\theta(s+t)}}{e^{-\theta t}} \\
&=e^{-\theta s} \\
&=P(X>s).
\end{align*}
Note that the event $X>s+t$ and $X>t$ implies and is implied by $X>s+t$ since $s>0$.



:::{.example}
Suppose the time $T$ between any two successive arrivals in a hospital emergency department 
has exponential distribution, $T \sim \operatorname{Exponential}(\lambda)$.
Historically, the mean of these inter-arrival times is 5 minutes. 
Estimate $\lambda$, and hence estimate

(i) $P(0<T< 5)$, 
(ii) $P(T<10 \mid T>5)$.

An estimate of $E(T)$ is 5. As $E(T)=\frac{1}{\lambda}$ we take 
$\frac{1}{5}$ as the estimate of $\lambda$.

(i) $P(0<T<5)=\int_0^5 \frac{1}{5} e^{-t / 5} d t=\left[-e^{-t / 5}\right]_0^5=1-e^{-1}=0.63212$.
(ii) We have \begin{align*}
P(T<10 \mid T>5) &=\frac{P(5<T<10)}{P(T>5)} \\
&=\frac{\int_5^{10} \frac{1}{5} e^{-t / 5} d t}{\int_5^{\infty} \frac{1}{5} e^{-t / 5} d t}=\frac{\left[-e^{-t / 5}\right]_5^{10}}{\left[-e^{-t / 5}\right]_5^{\infty}} \\
&=1-e^{-1}=0.63212 .
\end{align*}

:::

### Normal distribution

#### Definition

A random variable $X$ is said to have the normal distribution with parameters $\mu$ and $\sigma^2$ if it has pdf of the form
\begin{equation}
f(x)=\frac{1}{\sqrt{2 \pi \sigma^2}} \exp \left\{-\frac{(x-\mu)^2}{2 \sigma^2}\right\}, \; -\infty<x<\infty
(\#eq:normal-pdf)
\end{equation}
where $-\infty<\mu<\infty$ and $\sigma>0$ are two given constants. 
We write $X \sim N\left(\mu, \sigma^2\right)$.

We will show later that $E(X)=\mu$ and $\operatorname{Var}(X)=\sigma^2$.

The density (pdf) is much easier to remember and work with when the mean $\mu=0$ and variance $\sigma^2=1$. This special case is called the *standard* normal distribution. In this case, we simply write:
\[f(x)=\frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{x^2}{2}\right\}.\]


It is easy to see that $f(x)>0$ for all $x$. 
Next, we show $\int_{-\infty}^{\infty} f(x) d x=1$ or total probability equals 1, 
so that $f(x)$ defines a valid pdf:

\begin{align*}
\int_{-\infty}^{\infty} f(x) d x &=\int_{-\infty}^{\infty} \frac{1}{\sqrt{2 \pi \sigma^2}} \exp \left\{-\frac{(x-\mu)^2}{2 \sigma^2}\right\} d x \\
&=\frac{1}{\sqrt{2 \pi}} \int_{-\infty}^{\infty} \exp \left\{-\frac{z^2}{2}\right\} d z \quad \text { (substitute } z=\frac{x-\mu}{\sigma} \text { so that } d x=\sigma d z) \\
&=\frac{1}{\sqrt{2 \pi}} 2 \int_0^{\infty} \exp \left\{-\frac{z^2}{2}\right\} d z \quad \text { (since the integrand is an even function) } \\
&=\frac{1}{\sqrt{2 \pi}} 2 \int_0^{\infty} \exp \{-u\} \frac{d u}{\sqrt{2 u}}\text { (substitute } u=\frac{z^2}{2} \text { so that } z=\sqrt{2 u} \text { and } d z=\frac{d u}{\sqrt{2 u}}) \\
&=\frac{1}{2 \sqrt{\pi}} 2 \int_0^{\infty} u^{\frac{1}{2}-1} \exp \{-u\} d u \quad \text { (rearrange the terms) } \\
&=\frac{1}{\sqrt{\pi}} \Gamma\left(\frac{1}{2}\right) \quad \text { (recall the definition of the Gamma function) }\\
&=\frac{1}{\sqrt{\pi}} \sqrt{\pi}=1 \quad \text { as } \Gamma\left(\frac{1}{2}\right)=\sqrt{\pi}.
\end{align*}

#### Expectation and variance

We claimed that $E(X)=\mu$ and $\operatorname{Var}(X)=\sigma^2$, 
and now will prove these results.

To prove that  $E(X)=\mu$, we prove the following two results:

\begin{equation}
X \sim N\left(\mu, \sigma^2\right) \longleftrightarrow Z \equiv \frac{X-\mu}{\sigma} \sim N(0,1)
(\#eq:standardise-normal)
\end{equation}
\begin{equation}
E(Z)=0
(\#eq:std-normal-mean-zero)
\end{equation}

Then by the linearity of expectations, i.e. if $X=\mu+\sigma Z$ for constants $\mu$ and $\sigma$ then $E(X)=\mu+\sigma E(Z)=\mu$, the result follows. To prove \@ref(eq:standardise-normal),
we first calculate the cdf of $Z$, given by:
\begin{align*}
\Phi(z) &=P(Z \leq z) \\
&=P\left(\frac{X-\mu}{\sigma} \leq z\right) \\
&=P(X \leq \mu+z \sigma) \\
&=\int_{-\infty}^{\mu+z \sigma} \frac{1}{\sqrt{2 \pi \sigma^2}} \exp \left\{-\frac{(x-\mu)^2}{2 \sigma^2}\right\} d x \\
&=\int_{-\infty}^z \frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{u^2}{2}\right\} d u, \quad[u=(x-\mu) / \sigma]
\end{align*}
and so the pdf of $Z$ is
\[\frac{d \Phi(z)}{d z}=\frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{z^2}{2}\right\} \text { for }-\infty<z<\infty,\]
by the fundamental theorem of calculus. This proves that $Z \sim N(0,1)$. The converse is proved just by reversing the steps. Thus we have proved \@ref(eq:standardise-normal) above. We use the $\Phi(\cdot)$ notation to denote the cdf of the standard normal distribution. Now:
\begin{align*}
E(Z) &=\int_{-\infty}^{\infty} z f(z) d z \\
&=\int_{-\infty}^{\infty} z \frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{z^2}{2}\right\} d z \\
&=\frac{1}{\sqrt{2 \pi}} \times 0=0,
\end{align*}
since the integrand $g(z)=z \exp \left\{-\frac{z^2}{2}\right\}$ is an odd function, i.e. $g(z)=-g(-z)$; for an odd function $g(z), \int_{-a}^a g(z) d z=0$ for any $a$. Therefore we have also proved 
\@ref(eq:std-normal-mean-zero) and hence $E(X) = \mu$.

To prove $\operatorname{Var}(X)=\sigma^2$, we show that $\operatorname{Var}(Z)=1$ 
where $Z=\frac{X-\mu}{\sigma}$ and then claim that
$\operatorname{Var}(X)=\sigma^2 \operatorname{Var}(Z)=\sigma^2$ from our earlier result. 
Since $E(Z)=0, \operatorname{Var}(Z)=E\left(Z^2\right)$, which is
calculated below:
\begin{align*}
E\left(Z^2\right) &=\int_{-\infty}^{\infty} z^2 f(z) d z \\
&=\int_{-\infty}^{\infty} z^2 \frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{z^2}{2}\right\} d z \\
&=\frac{2}{\sqrt{2 \pi}} \int_0^{\infty} z^2 \exp \left\{-\frac{z^2}{2}\right\} d z \quad \text { (since the integrand is an even function) } \\
&=\frac{2}{\sqrt{2 \pi}} \int_0^{\infty} 2 u \exp \{-u\} \frac{d u}{\sqrt{2 u}} \quad \text { (substitute } u=\frac{z^2}{2} \text { so that } z=\sqrt{2 u} \text { and } d z=\frac{d u}{\sqrt{2 u}} ) \\
&=\frac{4}{2 \sqrt{\pi}} \int_0^{\infty} u^{\frac{1}{2}} \exp \{-u\} d u \\
&=\frac{2}{\sqrt{\pi}} \int_0^{\infty} u^{\frac{3}{2}}-1 \exp \{-u\} d u \\
&=\frac{2}{\sqrt{\pi}} \Gamma\left(\frac{3}{2}\right) \quad \text { (definition of the gamma function) }\\
&=\frac{2}{\sqrt{\pi}}\left(\frac{3}{2}-1\right) \Gamma\left(\frac{3}{2}-1\right) \quad \text { (reduction property of the gamma function) }\\
&=\frac{2}{\sqrt{\pi}} \frac{1}{2} \sqrt{\pi} \quad\text { (since } \Gamma\left(\frac{1}{2}\right) =\sqrt{\pi}) \\
&=1,
\end{align*}
as we hoped for! This proves $\operatorname{Var}(X)=\sigma^2$.

#### Linear transformation of a Normal random variable

::: {.theorem}
Suppose $X \sim N\left(\mu, \sigma^2\right)$ and $a$ and $b$ are constants. Then the distribution of $Y=a X+b$ is $N\left(a \mu+b, a^2 \sigma^2\right)$
:::

::: {.proof}
The result that $Y$ has mean $a \mu+b$ and variance $a^2 \sigma^2$ can already be claimed from the linearity of the expectations and the variance result for linear functions. What remains to be proved is that the normality of $Y$, i.e. how can we claim that $Y$ will follow the normal distribution too? For this, we note that
\[Y=a X+b=a(\mu+\sigma Z)+b=(a \mu+b)+a \sigma Z\]
since $X=\mu+\sigma Z$. Now we use \@ref(eq:standardise-normal) to claim the normality of $Y$.
:::

#### Standard normal distribution

Now we can claim that the normal pdf \@ref(eq:normal-pdf) 
is symmetric about the mean $\mu$. The spread of the
pdf is determined by $\sigma$, the standard deviation of the distribution. When $\mu = 0$ and 
$\sigma = 1$,
the normal distribution $N(0, 1)$ is called is called the standard normal distribution. 
We often use $Z$ to denote a random variable with standard normal distribution.
Next, we will show that the standard normal distribution can 
is used to calculate probabilities of interest for any normal 
distribution. 

Suppose $X \sim N\left(\mu, \sigma^2\right)$ and we are interested in finding $P(a \leq X \leq b)$ for two constants $a$ and $b$.
\begin{align*}
P(a \leq X \leq b) &=\int_a^b f(x) d x \\
&=\int_a^b \frac{1}{\sqrt{2 \pi \sigma^2}} \exp \left\{-\frac{(x-\mu)^2}{2 \sigma^2}\right\} d x \\
&\left.=\frac{1}{\sqrt{2 \pi}} \int_{\frac{a-\mu}{\sigma}}^\frac{b-\mu}{\sigma} \exp \left\{-\frac{z^2}{2}\right\} d z \quad \text { [substituted } z=\frac{x-\mu}{\sigma} \text { so that } d x=\sigma d z\right] \\
&=\int_{-\infty}^{\frac{b-\mu}{\sigma}} \frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{z^2}{2}\right\} d z-\int_{-\infty}^{\frac{a-\mu}{\sigma}} \frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{z^2}{2}\right\} d z \\
&=P\left(Z \leq \frac{b-\mu}{\sigma}\right)-P\left(Z \leq \frac{a-\mu}{\sigma}\right) \\
&=\text { cdf of } Z \text { at } \frac{b-\mu}{\sigma}-\text { cdf of } Z \text { at } \frac{a-\mu}{\sigma} \\
&=\Phi\left(\frac{b-\mu}{\sigma}\right)-\Phi\left(\frac{a-\mu}{\sigma}\right)
\end{align*}
where we use the notation $\Phi(\cdot)$ to denote the cdf of the standard normal
distribution, i.e.
\[\Phi(z)= P(Z \leq z)=\int_{-\infty}^z \frac{1}{\sqrt{2 \pi}} \exp \left\{-\frac{u^2}{2}\right\} d u.\]
This result allows us to find the probabilities about a normal random variable $X$ of any mean $\mu$ and variance $\sigma^2$ through the probabilities of the standard normal random variable $Z$. For this reason, only $\Phi(z)$ is tabulated. Further more, due to the symmetry of the pdf of $Z, \Phi(z)$ is tabulated only for positive $z$ values. Suppose $a>0$, then
\begin{align*}
\Phi(-a)=P(Z \leq-a) &=P(Z>a) \\
&=1-P(Z \leq a) \\
&=1-\Phi(a).
\end{align*}

In R, we use the function `pnorm` to calculate the probabilities.
`pnorm` which
has arguments `mean` (to specify $\mu$) and `sd` (to specify $\sigma$,
the standard deviation). By default `mean = 0` and `sd = 1`,
so by default `pnorm` calculates the standard normal cdf.
So, we use the command 
```{r}
pnorm(1)
```
to calculate $\Phi(1)=P(Z \leq 1)$. 
We can also use the command 
```{r}
pnorm(15, mean=10, sd=2)
```
to calculate $P(X \leq 15)$ when $X \sim N\left(\mu=10, \sigma^2=4\right)$ directly.

1. $P(-1<Z<1)=\Phi(1)-\Phi(-1)=0.6827$. This means that $68.27 \%$ of the probability lies within 1 standard deviation of the mean.
2. $P(-2<Z<2)=\Phi(2)-\Phi(-2)=0.9545$. This means that $95.45 \%$ of the probability lies within 2 standard deviations of the mean.
3. $P(-3<Z<3)=\Phi(3)-\Phi(-3)=0.9973$. This means that $99.73 \%$ of the probability lies within 3 standard deviations of the mean.

We are often interested in the quantiles (inverse-cdf of probability), $\Phi^{-1}(\cdot)$ of the normal distribution for various reasons. We find the $p$th quantile by issuing the R command `qnorm(p)`

1. $\texttt{qnorm(0.95)} = \Phi^{-1}(0.95) = 1.645$.
This means that the 95th percentile of the standard normal distribution is 1.645. This also means that $P(-1.645<Z<1.645)=\Phi(1.645)-$ $\Phi(-1.645)=0.90$.
2. $\texttt{qnorm(0.975)} = \Phi^{-1}(0.975) = 1.96$. This means that the 97.5th percentile of the standard normal distribution is 1.96. This also means that $P (-1.96 < Z < 1.96) = \Phi(1.96) -
\Phi(-1.96) = 0.95$.

:::{.example}
Suppose the marks in MATH1024 follow the normal distribution with mean
58 and standard deviation 32.25.

1. What percentage of students will fail (i.e. score less than 40) in MATH1024?   
Answer: $\texttt{pnorm(40, mean=58, sd=32.25)} = 28.84\%$.
2. What percentage of students will get an A result (score greater than 70)?    
Answer: $\texttt{1 - pnorm(70, mean=58, sd=32.25)} = 35.49\%$.
3. What is the probability that a randomly selected student will score more than 90?   
Answer: $\texttt{1 - pnorm(90, mean=58, sd=32.25)} = 0.1605$.
4. What is the probability that a randomly selected student will score less than 25?   
Answer: $\texttt{pnorm(25, mean=58, sd=32.25) = 0.1531}$. Ouch!
5. What is the probability that a randomly selected student scores a 2:1 (i.e. a mark between
60 and 70)? Left as an exercise.
:::


:::{.example}
A lecturer set and marked an examination and found that the distribution
of marks was $N (42, 14^2 )$. The school's policy is to present scaled marks whose distribution is
$N (50, 15^2 )$. What linear transformation should the lecturer apply to the raw marks to accomplish
this and what would the raw mark of 40 be transformed to?


Let $X$ be the raw mark and $Y$ the scaled mark. We have
$X \sim N (\mu_x = 42, \sigma_x^2 = 14^2 )$ and aim to define the scaling such that
$Y \sim N (\mu_y = 50, \sigma_y^2 = 15^2)$. 
If we standardise both variables, they should each have standard normal
distribution, so we choose $Y$ such that
\[\frac{X - \mu_x}{\sigma_x} = \frac{Y - \mu_y}{\sigma_y}\]
giving 
\[Y = \mu_y + \frac{\sigma_y}{\sigma_x}(X - \mu_x) = 50 + \frac{15}{14}(X - 42).\]

Now at raw mark $X = 40$, the transformed mark would be
\[Y = 50 + \frac{15}{14}(40 - 42) = 47.86.\]
:::

#### Log-normal distribution

If $X \sim N (\mu, \sigma^2)$ then the random variable $Y = exp(X)$ is called a 
log-normal random variable
and its distribution is called a log-normal distribution with parameters $\mu$ and $\sigma^2$.

The mean of the random variable $Y$ is given by
\begin{align*}
E(Y) &=E[\exp (X)] \\
&=\int_{-\infty}^{\infty} \exp (x) \frac{1}{\sigma \sqrt{2 \pi}} \exp \left\{-\frac{(x-\mu)^2}{2 \sigma^2}\right\} d x \\
&=\exp \left\{-\frac{\mu^2-\left(\mu+\sigma^2\right)^2}{2 \sigma^2}\right\} \int_{-\infty}^{\infty} \frac{1}{\sigma \sqrt{2 \pi}} \exp \left\{-\frac{x^2-2\left(\mu+\sigma^2\right) x+\left(\mu+\sigma^2\right)^2}{2 \sigma^2}\right\} d x \\
&=\exp \left\{-\frac{\mu^2-\left(\mu+\sigma^2\right)^2}{2 \sigma^2}\right\} \quad \text{(integrating a 
$N(\mu+\sigma^2, \sigma^2)$ random variable over its domain) }\\
&=\exp \left\{\mu+\sigma^2 / 2\right\}.
\end{align*}
Similarly, one can show that
\begin{align*}
E(Y^2) &=E[\exp (2 X)] \\
&=\int_{-\infty}^{\infty} \exp (2 x) \frac{1}{\sigma \sqrt{2 \pi}} \exp \left\{-\frac{(x-\mu)^2}{2 \sigma^2}\right\} d x \\
&=\cdots \\
&=\exp \left\{2 \mu+2 \sigma^2\right\}.
\end{align*}
Hence, the variance is given by
\[\operatorname{Var}(Y)=E(Y^2)-(E(Y))^2=\exp \left\{2 \mu+2 \sigma^2\right\}-\exp \left\{2 \mu+\sigma^2\right\}.\]

The log-normal
distribution is often used in practice for modelling economic variables of interest in business
and finance, e.g. volume of sales, income of individuals. You do not need to remember the mean
and variance of the log-normal distribution.

## Joint distributions

### Introduction 

Often we need to study more than one random variable, e.g. height and weight, simultaneously,
so that we can exploit the relationship between them to make inferences about their properties.
Multiple random variables are studied through their joint probability distribution. In this section
we will study covariance and correlation and then discuss when random variables are independent.

### Joint distribution of discrete random variables

If $X$ and $Y$ are discrete, the quantity $f(x, y) = P (X = x \cap Y = y)$ is called 
the joint probability
mass function (joint pmf) of $X$ and $Y$. To be a joint pmf, $f (x, y)$ needs to satisfy 
two conditions:
\[f (x, y) \geq 0 \quad \text{ for all $x$ and $y$}\] and
\[\sum_{\text{all $x$}} \sum_{\text{all $y$}} f(x, y) = 1.\]
The marginal probability mass functions (marginal pmfs) of $X$ and $Y$ are respectively
\[f_X(x) = \sum_y f(x, y), \quad f_Y(y) = \sum_x f(x, y)\]
Use the identity $\sum_{x} \sum_{y} f(x, y) = 1$
 to prove that $f_X(x)$ and $f_Y(y)$ are really pmfs.

:::{.example #joint-pdf-discrete}
Suppose that two fair dice are tossed independently one after the other. Let
\[X = \begin{cases}
-1 & \text{if the result from die 1 is larger} \\
0 & \text{if the results are equal} \\
1 & \text{if the result from die 1 is smaller}
\end{cases}
\]
and let $Y = |\text{difference between the two dice}|$.
Find the joint
probability pmf for $X$ and $Y$.

There are 36 possible outcomes for the results of the dice rolls,
and each gives a pair of values $(x, y)$ for $X$ and $Y$.

```{r, echo = FALSE}
x <- outer(1:6, 1:6, FUN = function(r1, r2){sign(r2 - r1)})
y <- outer(1:6, 1:6, FUN = function(r1, r2){abs(r2 - r1)})

x_y <- matrix(paste("(", x, ", ", y, ")", sep = ""), nrow = 6)
rownames(x_y) <- 1:6
colnames(x_y) <- 1:6

center_tab(x_y, row.names = TRUE, linesep = "")
```



The joint pmf is given in Table \@ref(tab:joint).

```{r, echo = FALSE, message = FALSE}
joint_counts <- tibble(x = as.numeric(x), y = as.numeric(y), count = 1) %>%
    group_by(x, y) %>%
    summarise(count = sum(count))
joint_counts_tab <- xtabs(count~x + y, joint_counts)
joint_prob_tab <- apply(joint_counts_tab, c(1, 2), function(x){paste("$\\frac{", x, "}{36}$", sep = "")})

library(huxtable)
as_hux(joint_prob_tab, add_rownames = "$x$", add_colnames = TRUE) %>%
    theme_basic() %>%
    set_tb_padding(1) %>%
    insert_row("", "$y$", "", "", "", "", "", after = 0) %>%
    merge_cells(1, 2:6) %>%
    set_align(1, everywhere, "center") %>%
    set_bold(2, everywhere, value = FALSE) %>%
    set_escape_contents(FALSE) %>%
    set_caption("\\label{tab:joint} The joint probabilities for $X$ and $Y$") %>%
    set_caption_pos("bottom")

```

```{r, include = FALSE}
cat(paste("\\frac{", joint_counts$count, "}{36} ", "& \\text{if $x = ", joint_counts$x, ", y = ", joint_counts$y, "$} \\\\ \n", sep = ""))

x_counts <- joint_counts %>%
    group_by(x) %>%
    summarise(count = sum(count))

cat(paste("\\frac{", x_counts$count, "}{36} ", "& \\text{if $x = ", x_counts$x, "$} \\\\ \n", sep = ""))

```

The marginal pmf for $X$ is
\[f_X(x) = \begin {cases}
  \frac{15}{36} & \text{if $x = -1$} \\ 
 \frac{6}{36} & \text{if $x = 0$} \\ 
 \frac{15}{36} & \text{if $x = 1$.} 
 \end{cases}\]
 
**Exercise**: Write down the marginal distribution of $Y$ and hence find the mean and variance of
$Y$.
:::

### Joint distribution of continuous random variables

If $X$ and $Y$ are continuous, a non-negative real-valued function $f(x, y)$ is called the joint probability density function (joint pdf) of $X$ and $Y$ if
\[\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f(x, y) d x \, d y=1\]
The marginal pdfs of $X$ and $Y$ are respectively
\[f_X(x)=\int_{-\infty}^{\infty} f(x, y) d y, \quad f_Y(y)=\int_{-\infty}^{\infty} f(x, y) d x\]

:::{.example #joint-pdf-cts}
Define a joint pdf by
\[f(x, y)= \begin{cases}
6 x y^2 & \text { if $0<x<1$ and $0<y<1$} \\
0 & \text { otherwise. }
\end{cases}\]
How can we show that the above is a pdf? It is non-negative for all $x$ and $y$ values. 
But does it integrate to 1? We are going to use the following rule.
:::

**Result**: Suppose that a real-valued function $f(x, y)$ is continuous in a region $A$,
where $A = \{(x, y) \text{ such that } a<x<b \text{ and } c<y<d\}$. Then
\[\int_A f(x, y) d x d y=\int_c^d \int_a^b f(x, y) d x \, dy.\]
The same result holds if $a$ and $b$ depend upon $y$,
but $c$ and $d$ should be free of $x$ and $y$. 
When we evaluate the inner integral $\int_a^b f(x, y) d x$, we treat $y$ as constant.

**Notes**: To evaluate a bivariate integral over a region $A$ we:

- Draw a picture of $A$ whenever possible.
- Rewrite the region $A$ as an intersection of two one-dimensional intervals. The first interval is obtained by treating one variable as constant.
- Perform two one-dimensional integrals.

:::{.example}
Continuing Example \@ref(exm:joint-pdf-cts),
\begin{align*}
\int_0^1 \int_0^1 f(x, y) d x d y &=\int_0^1 \int_0^1 6 x y^2 d x d y \\
&=6 \int_0^1 y^2 d y \int_0^1 x d x \\
&=3 \int_0^1 y^2 d y\left[\text { as } \int_0^1 x d x=\frac{1}{2}\right] \\
&=1 .\left[\text { as } \int_0^1 y^2 d y=\frac{1}{3}\right]
\end{align*}
Now we can find the marginal pdfs as well.
\[f_X(x)=2 x, \, 0<x<1, \quad f_Y(y)=3 y^2, \, 0<y<1.\]
:::

The probability of any event in the two-dimensional space can be found by integration 
and again more details will be provided in the second-year module
MATH2011, Statistical Distribution Theory. You will come across 
multivariate integrals in the second semester module MATH1060,
Multivariable Calculus. 

### Covariance and correlation

We first define the expectation of a real-valued scalar function $g(X, Y)$ of $X$ and $Y$ :
\[E[g(X, Y)]= \begin{cases}\sum_x \sum_y g(x, y) f(x, y) & \text { if } X \text { and } Y \text { are discrete } \\ \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} g(x, y) f(x, y) d x d y & \text { if } X \text { and } Y \text { are continuous. }\end{cases}\]

:::{.example}
Continuing Example \@ref(exm:joint-pdf-discrete),
let $g(x, y)=x y$.
\[E(X Y)=(-1)(0) 0+(-1)(1) \frac{5}{36}+\cdots+(1)(5) \frac{1}{36}=0.\]
**Exercises**: Try $g(x, y)=x$. It will be the same thing as $E(X)=\sum_x x f_X(x)$.
:::

We will not consider any continuous examples as the second-year module MATH2011 will study them in detail.

Suppose that two random variables $X$ and $Y$ have joint pmf or pdf $f(x, y)$ and let $E(X)=\mu_x$ and $E(Y)=\mu_y$. The covariance between $X$ and $Y$ is defined by
\[\operatorname{Cov}(X, Y)=E\left[\left(X-\mu_x\right)\left(Y-\mu_y\right)\right]=E(X Y)-\mu_x \mu_y.\]
Let $\sigma_x^2=\operatorname{Var}(X)=E\left(X^2\right)-\mu_x^2$ and $\sigma_y^2=\operatorname{Var}(Y)=E\left(Y^2\right)-\mu_y^2$. The correlation coefficient between $X$ and $Y$ is defined by:
\[\operatorname{Corr}(X, Y)=\frac{\operatorname{Cov}(X, Y)}{\sqrt{\operatorname{Var}(X) \operatorname{Var}(Y)}}=\frac{E(X Y)-\mu_x \mu_y}{\sigma_x \sigma_y}.\]
It can be proved that for any two random variables, $-1 \leq \operatorname{Corr}(X, Y) \leq 1$. 
The correlation $\operatorname{Corr}(X, Y)$ is a measure of linear dependency between two random 
variables $X$ and $Y$, and it is free of the measuring units of $X$ and $Y$ as the units cancel
in the ratio.


### Independence {#sec:indep}

Independence is an important concept. Recall that we say two events $A$ and $B$ are independent if $P(A \cap B)=P(A) \times P(B)$. We use the same idea here. Two random variables $X$ and $Y$ having the joint pdf or pmf $f(x, y)$ are said to be independent if and only if $f(x, y)=f_X(x) \times f_Y(y)$ for *all* $x$ and $y$.

In the discrete case $X$ and $Y$ are independent if each cell probability, $f(x, y)$, is the product of the corresponding row and column totals. In Example \@ref(exm:joint-pdf-discrete) 
$X$ and $Y$ are not independent. 

:::{.example}
Suppose $X$ and $Y$ have joint pdf given by the probability table:

<!-- \begin{tabular}{cc|ccc|c} -->
<!-- \multicolumn{2}{c}{} & & \multicolumn{3}{c}{$y$} \\ -->
<!-- \multicolumn{2}{c}{} & 1 & 2 & 3 & Total \\ -->
<!-- \hline \multirow{2}{*}{$x$} & 0 & $\frac{1}{6}$ & $\frac{1}{12}$ & $\frac{1}{12}$ & $\frac{1}{3}$ \\ -->
<!-- & 1 & $\frac{1}{4}$ & $\frac{1}{8}$ & $\frac{1}{8}$ & $\frac{1}{2}$ \\ -->
<!-- & 2 & $\frac{1}{12}$ & $\frac{1}{24}$ & $\frac{1}{24}$ & $\frac{1}{6}$ \\ -->
<!-- \hline & Total & $\frac{1}{2}$ & $\frac{1}{4}$ & $\frac{1}{4}$ & 1 -->
<!-- \end{tabular} -->

|  |  |  | $y$ |  |  |
| :---: | :---: | :---: | :---: | :---: | :---: |
|  |  | 1 | 2 | 3 | Total |
| $x$ | 0 | $\frac{1}{6}$ | $\frac{1}{12}$ | $\frac{1}{12}$ | $\frac{1}{3}$ |
|  | 1 | $\frac{1}{4}$ | $\frac{1}{8}$ | $\frac{1}{8}$ | $\frac{1}{2}$ |
|  | 2 | $\frac{1}{12}$ | $\frac{1}{24}$ | $\frac{1}{24}$ | $\frac{1}{6}$ |
|  | Total | $\frac{1}{2}$ | $\frac{1}{4}$ | $\frac{1}{4}$ | 1 |


Verify that in the following example $X$ and $Y$ are independent. 
We need to check all 9 cells.
:::

:::{.example}
Let $f(x, y)=6 x y^2, 0<x<1,0<y<1$. Check that $X$ and $Y$ are independent.
:::

:::{.example}
Let $f(x, y)=2 x, 0 \leq x \leq 1,0 \leq y \leq 1$. Check that $X$ and $Y$ are independent.
:::

Sometimes the joint pdf may look like something you can factorise, but $X$ and $Y$ may 
not be independent because they may be related in the domain. For instance,

1. $f(x, y)=\frac{21}{4} x^2 y, x^2 \leq y \leq 1$. Not independent!
2. $f(x, y)=e^{-y}, 0<x<y<\infty$. Not independent!

Here are some useful consequences of independence:

- Suppose that $X$ and $Y$ are independent random variables. Then
\[P(X \in A, Y \in B)=P(X \in A) \times P(Y \in B)\]
for any events $A$ and $B$. That is, the joint probability can be obtained as the product of the marginal probabilities. We will use this result in the next section. For example, suppose Jack and Jess are two randomly selected students. Let $X$ denote the height of Jack and $Y$ denote the height of Jess. Then
\[P(X<182 \text { and } Y>165)=P(X<182) \times P(Y>165) .\]
This is true for any numbers other than the example numbers 182 and 165, and for any inequalities.
- Let $g(x)$ be a function of $x$ only and $h(y)$ be a function of $y$ only. Then, if $X$ and $Y$ are independent,
\[E[g(X) h(Y)]=E[g(X)] \times E[h(Y)].\]
As a special case, let $g(x)=x$ and $h(y)=y$. Then we have
\[E(X Y)=E(X) \times E(Y).\]
Consequently, for independent random variables $X$ and $Y, \operatorname{Cov}(X, Y)=0$ and $\operatorname{Corr}(X, Y)=$ 0. But the converse is not true in general. That is, merely having $\operatorname{Corr}(X, Y)=0$ does not imply that $X$ and $Y$ are independent random variables.

<!-- IDEA: give an example of this, or set as an exercise? -->

## Sums of random variables {#sec:sum-rvs}

In this section we consider sums of random variables, which arise frequently in both practice and
theoretical results. For example, the mark achieved in an exam is the sum of the marks for each
question, and the sample mean is proportional to the sum of the sample values. 

Suppose we have obtained a random sample from a distribution with pmf or pdf $f(x)$, so that $X$ can either be a discrete or a continuous random variable. We will learn more about random sampling in the next chapter. Let $X_1, \ldots, X_n$ denote the random sample of size $n$ where $n$ is a positive integer. We use upper case letters since each member of the random sample is a random variable. For example, I toss a fair coin $n$ times and let $X_i$ take the value 1 if a head appears in the $i$ th trial and 0 otherwise. Now I have a random sample $X_1, \ldots, X_n$ from the Bernoulli distribution with probability of success equal to $0.5$ since the coin is assumed to be fair.

We can get a random sample from a continuous random variable as well. Suppose it is known that the distribution of the heights of first-year students is normal with mean 175 centimetres and standard deviation 8 centimetres. I can randomly select a number of first-year students and record each student's height.

Suppose $X_1, \ldots, X_n$ is a random sample from a population with distribution $f(x)$. Then it can be shown that the random variables $X_1, \ldots, X_n$ are mutually independent, i.e.
\[P\left(X_1 \in A_1, X_2 \in A_2, \ldots, X_n \in A_n\right)=P\left(X_1 \in A_1\right) \times P\left(X_2 \in A_2\right) \times \cdots P\left(X_n \in A_n\right)\]
for any set of events, $A_1, A_2, \ldots A_n$. That is, the joint probability can be obtained as the product of individual probabilities. An example of this for $n=2$ was given in Section \@ref(sec:indep).

::: {.example name="Distribution of the sum of independent binomial random variables"}
Suppose $X \sim \operatorname{Bin}(m, p)$ and $Y \sim \operatorname{Bin}(n, p)$ independently. Note that $p$ is the same in both distributions. Using the above fact that joint probability is the multiplication of individual probabilities, we can conclude that $Z=X+Y$ has the binomial distribution. It is intuitively clear that this should happen since $X$ comes from $m$ Bernoulli trials and $Y$ comes from $n$ Bernoulli trials independently, so $Z$ comes from $m+n$ Bernoulli trials with common success probability $p$. 


Next we will prove the result mathematically, by finding the probability mass function of $Z=X+Y$ directly and observing that it is of the appropriate form. 
In our proof, we will need to use the fact that
\[\sum_{x+y=z} \binom{m}{x}\binom{n}{y} = \binom{m+n}{z}\]
  where the above sum is also over all possible integer values of $x$ and $y$ such that $0 \leq x \leq m$ and $0 \leq y \leq n$. This fact may be proved by using the binomial theorem,
  but we state it here without proof.

Note that
\[P(Z=z)=P(X=x, Y=y)\]
subject to the constraint that $x+y=z$, $0 \leq x \leq m$, $0 \leq y \leq n$. Thus,
\begin{align*}
P(Z=z) &= \sum_{x+y=z} P(X=x, Y=y) \\
&= \sum_{x+y=z} \binom{m}{x} p^x(1-p)^{m-x}\binom{n}{y} p^y(1-p)^{n-y} \\
&= \sum_{x+y=z} \binom{m}{x}\binom{n}{y} p^z(1-p)^{m+n-z} \\
&= p^z(1-p)^{m+n-z} \sum_{x+y=z} \binom{m}{x}\binom{n}{y} \\
&= \binom{m+n}{z} p^z(1-p)^{m+n-z}, \; \; \text{using the fact above.}
\end{align*}
Thus, we have proved that the sum of independent binomial random variables with common probability is binomial as well. This is called the reproductive property of random variables. 
:::

Now we will state two main results without proof. The proofs will presented in the second-year distribution theory module MATH2011. Suppose that $X_1, \ldots, X_n$ is a random sample from a population distribution with finite variance, and suppose that $E\left(X_i\right)=\mu_i$ and $\operatorname{Var}\left(X_i\right)=\sigma_i^2$. Define a new random variable
\[Y=a_1 X_1+a_2 X_2+\cdots+a_n X_n\]
where $a_1, a_2, \ldots, a_n$ are constants. Then:

1. $E(Y)=a_1 \mu_1+a_2 \mu_2+\cdots+a_n \mu_n$.
2. $\operatorname{Var}(Y)=a_1^2 \sigma_1^2+a_2^2 \sigma_2^2+\cdots+a_n^2 \sigma_n^2$.

For example, if $a_i=1$ for all $i=1, \ldots, n$, the two results above imply that:

- The expectation of the sum of independent random variables is the sum of the expectations of the individual random variables
- the variance of the sum of independent random variables is the sum of the variances of the individual random variables.

The second result is *only* true for independent random variables, e.g. random samples. Now we will consider many examples.

:::{.example name="Mean and variance of binomial distribution"}
Suppose $Y \sim \operatorname{Bin}(n, p)$. Then we can write:
\[Y=X_1+X_2+\ldots+X_n\]
where each $X_i$ is an independent Bernoulli trial with success probability $p$. We have shown before that, $E\left(X_i\right)=p$ and $\operatorname{Var}\left(X_i\right)=p(1-p)$ by direct calculation. Now the above two results imply that:
\[
E(Y)=E\left(\sum_{i=1}^n X_i\right)=p+p+\ldots+p=n p . \\
\operatorname{Var}(Y)=\operatorname{Var}\left(X_1\right)+\cdots+\operatorname{Var}\left(X_n\right)=p(1-p)+\ldots+p(1-p)=n p(1-p) .
\]
Thus we avoided the complicated sums used to derive $E(X)$ and $\operatorname{Var}(X)$ in 
Sections \@ref(sec:bin-mean) and \@ref(sec:bin-var).
:::

:::{.example name="Mean and variance of negative binomial distribution"}
Recall that the negative binomial random variable $Y$ is the number of trials needed to obtain the
$r$th success in a sequence of independent Bernoulli trials, each with success probability $p$.
Let $X_i$ be the number of trials needed after the $(i-1)$th success to obtain the $i$th success.
Each $X_i$ is a geometric random variable and $Y=X_1+\cdots+X_r$. Hence
\[E(Y)=E(X_1)+\cdots+E(X_r)=1 / p+\cdots+1 / p=r / p\]
and
\[\operatorname{Var}(Y)=\operatorname{Var}(X_)+\cdots+\operatorname{Var}(X_r)=(1-p) / p^2+\cdots+(1-p) / p^2=r(1-p) / p^2.\]
:::

:::{.example name="Sum of independent Normal random variables"}
Suppose that $X_i \sim N\left(\mu_i, \sigma_i^2\right), i=1,2, \ldots, k$ 
are independent random variables. Let $a_1, a_2, \ldots, a_k$ be constants 
and suppose that
\[Y=a_1 X_1+\cdots+a_k X_k.\]
Then we can prove that:
\[Y \sim N\left(\sum_{i=1}^k a_i \mu_i, \sum_{i=1}^k a_i^2 \sigma_i^2\right).\]
It is clear that $E(Y)=\sum_{i=1}^k a_i \mu_i$ and $\operatorname{Var}(Y)=\sum_{i=1}^k a_i^2 \sigma_i^2$. But that $Y$ has the normal distribution cannot yet be proved with the theory we know.
This proof will be provided in the second-year distribution theory module MATH2011.

As a consequence of the stated result we can easily see the following. 
Suppose $X_1$ and $X_2$ are independent $N\left(\mu, \sigma^2\right)$ random variables. 
Then $2 X_1 \sim N\left(2 \mu, 4 \sigma^2\right), X_1+X_2 \sim N\left(2 \mu, 2 \sigma^2\right)$, 
and $X_1-X_2 \sim N\left(0,2 \sigma^2\right)$. Note that $2 X_1$ and $X_1+X_2$ have different 
distributions.
Suppose that $X_i \sim N\left(\mu, \sigma^2\right), i=1, \ldots, n$ are independent. Then
\[X_1+\cdots+X_n \sim N\left(n \mu, n \sigma^2\right),\]
and consequently
\[\bar{X}=\frac{1}{n}\left(X_1+\cdots+X_n\right) \sim N\left(\mu, \frac{\sigma^2}{n}\right).\]
:::


## The Central Limit Theorem {#sec:clt}

### Introduction

The sum (and average) of independent random variables show a remarkable behaviour in practice which is captured by the Central Limit Theorem (CLT). These random variables do not even have to be continuous, all we require is that they are independent and each of them has a finite mean and a finite variance. A version of the CLT follows.

### Statement of the Central Limit Theorem (CLT)

Let $X_{1}, \ldots, X_{n}$ be independent random variables with finite $E\left(X_{i}\right)=\mu_{i}$ and finite $\operatorname{Var}\left(X_{i}\right)=\sigma_{i}^{2}$. Define $Y=\sum_{i=1}^{n} X_{i}$. Then, for a sufficiently large $n$, the central limit theorem states that $Y$ is approximately normally distributed with
\[E(Y)=\sum_{i=1}^{n} \mu_{i}, \quad \operatorname{Var}(Y)=\sum_{i=1}^{n} \sigma_{i}^{2}.\]
This also implies that $\bar{X}=\frac{1}{n} Y$ also follows the normal distribution approximately,
as the sample size $n \rightarrow \infty$. In particular, if $\mu_{i}=\mu$ and
$\sigma_{i}^{2}=\sigma^{2}$, i.e. all means are equal and all variances are equal, then the CLT
states that, as $n \rightarrow \infty$,
\[\bar{X} \sim N\left(\mu, \frac{\sigma^{2}}{n}\right).\]
Equivalently,
\[\frac{\sqrt{n}(\bar{X}-\mu)}{\sigma} \sim N(0,1)\]
as $n \rightarrow \infty$. The notion of convergence is explained by the convergence of distribution of $\bar{X}$ to that of the normal distribution with the appropriate mean and variance. It means that the $\mathrm{cdf}$ of the left hand side, $\sqrt{n} \frac{(\bar{X}-\mu)}{\sigma}$, converges to the cdf of the standard normal random variable, $\Phi(\cdot)$. In other words,
\[\lim _{n \rightarrow \infty} P\left(\sqrt{n} \frac{(\bar{X}-\mu)}{\sigma} \leq z\right)=\Phi(z), \quad-\infty<z<\infty.\]
So for "large samples", we can use $N(0,1)$ as an approximation to the sampling distribution of $\sqrt{n}(\bar{X}-\mu) / \sigma$. This result is 'exact', i.e. no approximation is required, if the distribution of the $X_{i}$ 's are normal in the first place --- this was discussed in the previous lecture.

How large does $n$ have to be before this approximation becomes usable? There is no definitive answer to this, as it depends on how "close to normal" the distribution of $X$ is. However, it is often a pretty good approximation for sample sizes as small as 20, or even smaller. It also depends on the skewness of the distribution of $X$; if the $X$-variables are highly skewed, then $n$ will usually need to be larger than for corresponding symmetric $X$-variables for the approximation to be good. 
We investigate in one example in Figure \@ref(fig:clt-demo).


```{r clt-demo, echo = FALSE, fig.cap = "Distribution of normalised sample means for samples of different sizes. Initially very skew (original distribution, $n=1$) becoming rapidly closer to standard normal (dashed line) with increasing $n$.", fig.width = 5, fig.height = 3, out.width = "60%", fig.align = "center"}
f <- function(z, n) {
    dgamma(z + sqrt(n), shape = n, scale = 1 / sqrt(n))
}

plot_data <- crossing(x = seq(-5, 5, length.out = 1000),
                      n = c(1, 2, 5, 10, 20, 50)) %>%
    mutate(pdf = f(x, n),
           n = as.factor(n))

plot_data %>%
    ggplot(aes(x = x, y = pdf, colour = n)) +
    geom_line() +
    geom_function(fun = dnorm, linetype = "dashed", colour = "black")
        
```

### Application of CLT to binomial distribution

We know that a binomial random variable $Y$ with parameters $n$ and $p$ is the number of successes in a set of $n$ independent Bernoulli trials, each with success probability $p$. We may write
\[Y=X_{1}+X_{2}+\cdots+X_{n},\]
where $X_{1}, \ldots, X_{n}$ are independent Bernoulli random variables with success probability $p$. It follows from the CLT that, for a sufficiently large $n, Y$ is approximately normally distributed with expectation $E(Y)=n p$ and variance $\operatorname{Var}(Y)=n p(1-p)$.

<!-- IDEA: generate plot? Really histograms? But then what bin width? -->
<!-- Maybe replace with CDF plots? -->
<!-- "Histograms of normalised sample means for Bernoulli $(p=0.8)$ samples of different sizes, converging to standard normal." -->

Hence, for given integers $y_{1}$ and $y_{2}$ between 0 and $n$ and a suitably large $n$, we have
\begin{align*}
P\left(y_{1} \leq Y \leq y_{2}\right) &=P\left\{\frac{y_{1}-n p}{\sqrt{n p(1-p)}} \leq \frac{Y-n p}{\sqrt{n p(1-p)}} \leq \frac{y_{2}-n p}{\sqrt{n p(1-p)}}\right\} \\
& \approx P\left\{\frac{y_{1}-n p}{\sqrt{n p(1-p)}} \leq Z \leq \frac{y_{2}-n p}{\sqrt{n p(1-p)}}\right\}
\end{align*}
where $Z \sim N(0,1)$.

We should take account of the fact that the binomial random variable $Y$ is integer-valued, and so 
$P(y_{1} \leq Y \leq y_{2})=P(y_{1}-f_{1} \leq Y \leq y_{2}+f_{2})$ for any two fractions
$0<f_{1}, f_{2}<1$. This is called
continuity correction and we take $f_{1}=f_{2}=0.5$ in practice.
\begin{align*}
P\left(y_{1} \leq Y \leq y_{2}\right) &=P\left(y_{1}-0.5 \leq Y \leq y_{2}+0.5\right) \\
&=P\left\{\frac{y_{1}-0.5-n p}{\sqrt{n p(1-p)}} \leq \frac{Y-n p}{\sqrt{n p(1-p)}} \leq \frac{y_{2}+0.5-n p}{\sqrt{n p(1-p)}}\right\} \\
& \approx P\left\{\frac{y_{1}-0.5-n p}{\sqrt{n p(1-p)}} \leq Z \leq \frac{y_{2}+0.5-n p}{\sqrt{n p(1-p)}}\right\} .
\end{align*}
What do we mean by a suitably large $n$? A commonly-used guideline is that the approximation is adequate if $n p \geq 5$ and $n(1-p) \geq 5$.

:::{.example}
A producer of natural yoghurt believed that the market share of their brand was $10 \%$. To investigate this, a survey of 2500 yoghurt consumers was carried out. It was observed that only 205 of the people surveyed expressed a preference for their brand. Should the producer be concerned that they might be losing market share?

Assume that the conjecture about market share is true. Then the number of people $Y$ who prefer this product follows a binomial distribution with $p=0.1$ and $n=2500$. So the mean is $n p=250$, 
the variance is $n p(1-p)=225$, and the standard deviation is 15 . The exact probability of 
observing $(Y \leq 205)$ is given by the sum of the binomial probabilities up to and including 205,
which is difficult to compute. However, this can be approximated by using the CLT:
\begin{align*}
P(Y \leq 205) &=P(Y \leq 205.5) \\
&=P\left\{\frac{Y-n p}{\sqrt{n p(1-p)}} \leq \frac{205.5-n p}{\sqrt{n p(1-p)}}\right\} \\
& \approx P\left\{Z \leq \frac{205.5-n p}{\sqrt{n p(1-p)}}\right\} \\
&=P\left\{Z \leq \frac{205.5-250}{15}\right\} \\
&=\Phi(-2.967)=0.0015.
\end{align*}
This probability is so small that it casts doubt on the validity of the assumption that the market share is $10 \%$.

Although the exact binomial probabilities are difficult to compute by hand,
in this case we may compute them in R. Recall $Y \sim \operatorname{Bin}(n = 2500, p = 0.1)$,
so $P(Y \leq 205)$ is
```{r}
pbinom(205, size = 2500, prob = 0.1)
```
In this case the normal approximation was good enough to correctly conclude that this
probability is very small (of the order of $0.1\%$), which was all we needed to answer 
the question of interest here.
:::

\clearpage
